Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-04 doc/JFLA2022/paper.tex
Macros for Hydras & co and Coq-community.

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

@@

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library on discrete math, written for the Coq proof assistant, and developed under the \emph{Coq-community} project. The proof scripts are
  accompanied with an electronic book, made with the help of the \emph{Alectryon} litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several Coq-community projects it requires and evolutions of Coq, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\subsection{Vision}
The \emph{Hydra-battles} project, part of \emph{Coq-community}~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \emph{Coq-community} is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the Coq-community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

\section{Integration within Coq-community}

@@

\emph{Coq-community} is an informal organization run by volunteer Coq users with objectives to maintain interesting Coq projects in the long-term and help Coq users to collaborate on documentation, tooling, etc.

@@

In the case of Coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%
\emph{Coq-community} makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \emph{Coq-community} hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the Coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new Coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \emph{Coq-community}.

\subsection{Interactions with other Coq-community projects}

Given the objectives of \emph{Coq-community}, it makes sense to propose a transfer anytime we encounter an interesting project that is insufficiently maintained.
%
After the project is adopted, large changes and refactorings are explicitly allowed.
%
This means, for instance, that we can consolidate libraries together or split them into several packages if it makes sense.

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-community}.

@@

The hydra-battles book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the Coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

@@

We solved this maintenance issue by moving to Alectryon, a tool to automatically record Coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by Alectryon from our Coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of Coq or versions of hydra-battles, we can easily spot unexpected changes).

@@

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as Alectryon).

@@

\emph{Hydra-battles} wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the Coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \textit{Coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

 \emph{Hydra-battles} is not limited to the study of ordinal numbers and applications.
 For instance, we are also developing 
a module about efficient exponentiation algorithms, and hope
to extend our project to new topics.

>>>>

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\Hydras}{Hydras \& Co. }
\newcommand{\Coqcommunity}{\emph{Coq-community} }

@@

\begin{abstract}
  \Hydras is a collaborative library on discrete math, written for the Coq proof assistant, and developed under the \Coqcommunity project. The proof scripts are
  accompanied with an electronic book, made with the help of the \emph{Alectryon} litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several \Coqcommunity projects it requires and evolutions of Coq, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\subsection{Vision}
The \Hydras project, part of \Coqcommunity~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \Coqcommunity is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \Hydras project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the \Coqcommunity project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

\section{Integration within \Coqcommunity}

@@

\Coqcommunity is an informal organization run by volunteer Coq users with objectives to maintain interesting Coq projects in the long-term and help Coq users to collaborate on documentation, tooling, etc.

@@

In the case of Coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%
\Coqcommunity makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \Coqcommunity hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the Coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new Coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \Coqcommunity.

\subsection{Interactions with other \Coqcommunity projects}

Given the objectives of \Coqcommunity, it makes sense to propose a transfer anytime we encounter an interesting project that is insufficiently maintained.
%
After the project is adopted, large changes and refactorings are explicitly allowed.
%
This means, for instance, that we can consolidate libraries together or split them into several packages if it makes sense.

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \Coqcommunity.

@@

The \Hydras book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the Coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

@@

We solved this maintenance issue by moving to Alectryon, a tool to automatically record Coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by Alectryon from our Coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of Coq or versions of \Hydras, we can easily spot unexpected changes).

@@

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was geared towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as Alectryon).

@@

\Hydras wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the Coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \Coqcommunity contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \textit{Coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

 \Hydras is not limited to the study of ordinal numbers and applications.
 For instance, we are also developing 
a module about efficient exponentiation algorithms, and hope
to extend our project to new topics.
